\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{fontenc}
\usepackage{inputenc}
\usepackage[pdftex,colorlinks,linkcolor=blue,citecolor=OliveGreen,urlcolor=blue,filecolor=blue]{hyperref}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage[dvipsnames]{xcolor}
\usepackage{booktabs}
\usepackage[numbers]{natbib}

\usepackage{amsmath,amsthm,amssymb}
\usepackage{graphicx}

\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{1em}{}
\titlespacing{\section}{0pt}{8pt}{4pt}
\titlespacing{\subsection}{0pt}{5pt}{1pt}
\titlespacing{\subsubsection}{0pt}{3pt}{0pt}

\begin{document}

\noindent{\bf Local and Latent: Dual Views of Symmetry Breaking in Cooperative Multi-Robot RL}

\noindent {\bf PI:} Qi Zhang, Assistant Professor, Computer Science Department, Worcester Polytechnic Institute
\noindent {\bf Cash funding needed:} \$42,450 \qquad
{\bf AWS Promotional Credits needed:} \$50,000\\
\noindent {\bf Amazon contacts:}
% Dingyang Chen $\langle$dingerwa@amazon.com$\rangle$,
Zhenyu Zhang $\langle$zhenyuzh@amazon.com$\rangle$,
Özalp Özer $\langle$ozalpo@amazon.com$\rangle$

\noindent{\bf Abstract }
Spatial symmetry is among the most effective inductive biases for robot learning, but in real multi-robot deployments it almost never holds exactly. Violations come in two structurally distinct forms: \emph{local} breakers, visible in the current state but confined to small regions (e.g., fixed obstacles), and \emph{latent} breakers, hidden from observation but fixed across an episode (e.g., wind or gravity). The two are orthogonal—state-dependent gating cannot find an unobserved breaker, and hidden-context inference cannot represent localized agent-dependent violations—yet realistic deployments exhibit both. We propose a dual-view theory and unified architecture. View~1 formalizes a locally factored partially symmetric Markov game and shows that critic-level symmetry error is controlled by discounted future breaker exposure, motivating a TD-learned per-agent gated critic. View~2 formalizes a latent symmetry-breaker POMDP and bounds value error by breaker-estimation error. The unified architecture pairs an equivariant latent encoder with per-agent gating, yielding testable theoretical bounds, a practical multi-robot RL algorithm, and benchmarks that isolate local, latent, and mixed symmetry-breaking regimes.\\
\noindent{\bf Keywords}
cooperative multi-robot RL, group equivariance, symmetry breaking


\section{Introduction}
Group symmetries—geometric transformations leaving system properties invariant—provide one of the most effective inductive biases in robot learning, with consistent gains in imitation learning~\citep{zeng2021transporter,wang2024equivariantdiffusion,seo2025survey} and reinforcement learning (RL)~\citep{vanderpol2020mdp,wang2022so2,luo2024reinforcement}, where exploration is expensive and demonstrations are scarce.
Formally, given a group $G$ acting on the states and actions of an MDP, the MDP is $G$-symmetric if its reward and transition are invariant under the group action.
The notion extends to multi-robot settings by lifting actions and invariance to joint spaces, including PI's prior work~\citep{chen2022communicationefficient,chen20243}, where the relevant group factors as $G = G_{\text{spatial}} \times \mathrm{S}_n$. We focus on the spatial component and leave permutation symmetry ($\mathrm{S}_n$) for future work.

In practice, symmetry almost never holds exactly. We observe that violations fall into two structurally distinct modes, separated by a single diagnostic: \emph{is the breaker present in an agent's current observation, or hidden from it?} \emph{Local} breakers are observable and confined to small regions of the per-agent state-action space, e.g., for warehouse robots near a fixed shelving unit, symmetry breaks only in the unit's vicinity, and at any moment only a subset of agents is affected. \emph{Latent} breakers are hidden global variables fixed within an episode, e.g., wind on aerial drones, gravity on inclined multi-arm setups, current on underwater robots, where dynamics would be exactly symmetric \emph{if} the vector were rotated together with state and action, yet the vector is not directly observable.

The two modes are structurally orthogonal: local methods cannot detect hidden breakers; a global latent variable is a poor inductive bias for representing agent-dependent localized violations. Realistic deployments exhibit both. This \emph{local-latent} dichotomy separates our proposal from prior work, which either treats single-agent settings where the dichotomy does not arise or globally relaxes symmetry without distinguishing structurally different modes~\citep{finzi2021rpp,wang2022approx,romero2022partial,vanderouderaa2022relaxing,hofgard2024relaxed,park2025approximate,chang2026partially}. Closing this gap is the goal of this proposal.

\section{Methods}
We consider a discounted cooperative Markov game with agent set $\mathcal{N} = \{1, \dots, N\}$,
global state space $\mathcal{S}=\mathcal{S}_1 \times \cdots \times \mathcal{S}_N$ decomposed into per-agent local states,
joint action space $\mathcal{A} = \mathcal{A}_1 \times \cdots \times \mathcal{A}_N$,
and team reward $r\colon \mathcal{S} \times \mathcal{A} \to \mathbb{R}$, 
equipped with a spatial group $G$ lifted to each agent's local state and action via group actions. We pursue two complementary views of why $G$ may fail to be exact: View~1 attributes failures to localized regions of the observed local state-action space, motivating a state-dependent gated critic; View~2 attributes them to an unobserved global breaker variable (wind, gravity, current) and recovers equivariance in the augmented state through breaker inference. The two will later be combined in a single architecture.

\subsection{View~1: symmetry breakers as locally observed state-action regions}
We posit a \emph{locally factored} partially symmetric Markov game in which the reward and transition decompose agent-wise:
$
r(s,a) = \sum_{i=1}^N r_i(s_i, a_i),
P(\cdot | s, a) = \prod_{i=1}^N P_i(\cdot | s_i, a_i)
$.
Each local factor carries its own finite symmetry group $G$ acting on $(\mathcal{S}_i, \mathcal{A}_i)$ and its own \emph{local breaker region} $\mathcal{B}_i \subseteq \mathcal{S}_i \times \mathcal{A}_i$. On the safe set $(s_i, a_i) \notin \mathcal{B}_i$ the local reward and transition are $G$-invariant; on $\mathcal{B}_i$ they may be arbitrary. The breaker region is itself $G$-invariant, so the safe/broken designation is a property of the orbit. This is more general than a single global gate over the joint state: symmetry can be exploited at agents whose local context is symmetric even when others' are not.
We introduce a locally routed model with routing parameter $\lambda_i(s_i,a_i) \in [0,1]$. Let $\overline r_i$ and $\overline P_i$ denote the $G$-orbit-averaged reward and transition. The routed factor interpolates between the equivariant and raw models: $\widetilde r_i^\lambda=(1-\lambda_i)\overline r_i+\lambda_i r_i$ and $\widetilde P_i^\lambda=(1-\lambda_i)\overline P_i+\lambda_i P_i$. Thus $\lambda_i=0$ recovers the fully equivariant model, while $\lambda_i=1$ recovers the raw model. The oracle breaker route $\lambda_i^*(s_i,a_i):=\mathbf{1}\{(s_i,a_i)\in\mathcal B_i\}$ is useful as a diagnostic, but our analysis and algorithm allow arbitrary learned routes.

\noindent{\bf Theory.}
For any bounded local critic $Q_i$, the routed model has a one-step Bellman error bounded by the residual of the equivariant approximation, scaled by the amount of averaging used, i.e., $|(\widetilde{\mathcal T}_i^\lambda Q_i)(s_i,a_i)-({\mathcal T}_i Q_i)(s_i,a_i)| \lesssim (1-\lambda_i(s_i,a_i))(\Delta_i^r+\gamma V_{i,\max}\Delta_i^P)$. Thus small $\lambda_i$ exploits equivariance when local residuals are small, while large $\lambda_i$ avoids bias near local breakers. We will use this one-step bound to show that local routing controls the induced joint-critic and policy-gradient errors without requiring a single global gate.
At the critic level, the key point is that a locally safe pair $(s_i,a_i)\notin\mathcal B_i$ can still have asymmetric value if future trajectories enter a breaker region. We will formalize this through a discounted future-exposure term $\beta_i^{\pi_i}(s_i,a_i)$ and show that critic asymmetry is controlled by this exposure, motivating gates learned from TD error rather than hand-coded breaker indicators.
Thus, the learned gate should not simply detect whether the current local state-action pair lies in a breaker region; it should estimate whether the current pair is likely to lead to asymmetric future values.



\noindent{\bf Algorithm.}
Algorithmically, this view leads to a per-agent routed critic. Each agent \(i\) uses a gate
\(\lambda_i(s_i,a_i)\in[0,1]\) to interpolate between an equivariant local critic and an unconstrained local critic:
\(Q_i^{\lambda}=(1-\lambda_i)Q_i^{\mathrm{eq}}+\lambda_i Q_i^{\mathrm{free}}\).
We will learn \(\lambda_i\) from TD error and counterfactual consistency signals, allowing each agent to exploit symmetry in locally safe regimes while routing around observable symmetry breakers.

\subsection{View~2: symmetry breakers as partially observed latent state}
For hidden-context breakers, the agent-wise local-region view does not fit: the breaker is a single global object (e.g., a wind or gravity vector) shared by all agents and fixed across the episode. Throughout this subsection we therefore use single-agent-style notation $(s, a, o, b)$, with the understanding that joint quantities decompose per-agent and $G$ acts simultaneously on all agents' frames; agent-wise factorization returns in \S\ref{sec:unified}. We model the environment as a \emph{latent symmetry-breaker POMDP} in which breaker $b \in \mathcal{B}$ is sampled at the start of each episode and held fixed; the agent receives observations $o_t \sim \Omega_b(\cdot | s_t)$ and history $h_t$ but never sees $b$.
The structural assumption is \emph{augmented equivariance}: for every $g \in G$,
$
r_{g\cdot b}(g \cdot s, g \cdot a) = r_b(s,a), \quad
P_{g\cdot b}(g \cdot C | g \cdot s, g \cdot a) = P_b(C | s,a), \quad
\Omega_{g\cdot b}(g \cdot D | g \cdot s) = \Omega_b(D | s)
$.
The system is symmetric if and only if $b$ is transformed alongside $(s,a)$; the visible state alone admits no group action consistent with the dynamics.

\noindent{\bf Theory.}
We will establish two key results that drive the algorithm. 
First, under Lipschitz assumptions $|r_b - r_{b'}| \le L_r d_\mathcal{B}(b, b')$ and $\|P_b - P_{b'}\|_{\text{TV}} \le L_p d_\mathcal{B}(b, b')$, conditioning on an estimate $\widehat{b}$ rather than the true $b$ yields a fixed-point value error bounded by $\tfrac{L_r + \gamma V_{\max} L_p}{1-\gamma}\,d_\mathcal{B}(b, \widehat{b})$, motivating auxiliary objectives that drive $d_\mathcal{B}(b, \widehat{b})$ down. 
Second, under augmented equivariance and an equivariant prior $\rho_0$ over $\mathcal{B}$, the Bayesian posterior over $b$ transforms equivariantly: $\beta^{g \cdot h_t}_t(g \cdot B) = \beta^{h_t}_t(B)$; the encoder $f_\eta$ should therefore be built so its output transforms in the same representation as $b$.

\noindent{\bf Algorithm.}
The architecture has three components. A latent breaker encoder $z_t = f_\eta(h_t)$ (recurrent or contrastive) is built so $f_\eta(g \cdot h_t) = g \cdot f_\eta(h_t)$, i.e., its output lives in the same representation as $b$. A breaker-conditioned policy and critic are equivariant in the augmented pair $(o, z)$: $\pi_\theta(g \cdot a | g \cdot o, g \cdot z) = \pi_\theta(a | o, z)$ and $Q_\omega(g \cdot o, g \cdot a, g \cdot z) = Q_\omega(o,a,z)$. The encoder is trained with auxiliary objectives in addition to TD: a dynamics loss $\mathcal{L}_{\text{dyn}} = \mathbb{E}[\ell(D_\psi(o_t, a_t, z_t), o_{t+1})]$, a contrastive loss pulling same-episode representations together (and pushing across-episode pairs apart), and an optional supervised loss $\mathcal{L}_{\text{sup}} = \mathbb{E}[d_\mathcal{B}(f_\eta(h_t), b)^2]$ when simulator labels are available. We will explore three variants: a point-estimate encoder producing $\widehat{b}_t \in \mathcal{B}$; a belief representation outputting distributional parameters; and canonicalization, rotating inputs into a frame aligned with $\widehat{b}_t$, applying a standard network, and rotating outputs back.

\subsection{Unifying perspective and combined architecture}
\label{sec:unified}
The two views compose because they operate at different granularities:
View 1 is per-agent and state-dependent, while View 2 is global and
per-episode. We therefore replace View 1's orbit-averaged branch with
View 2's breaker-conditioned equivariant branch, using a shared latent
encoder \(z_t=f_\eta(h_t)\) that is broadcast to all agents, while retaining the
per-agent gate. Each agent's critic is
$
Q_i(s_i,a_i,h_t)
=
\bigl(1-\lambda_{i,\psi}(s_i,a_i)\bigr)q^{\mathrm{cond}}_{i,\theta}(s_i,a_i,z_t)
+
\lambda_{i,\psi}(s_i,a_i)
q^{\mathrm{raw}}_{i,\phi}(s_i,a_i,h_t)
$,
and the joint critic is
$
Q(s,a,h_t)=\sum_i Q_i(s_i,a_i,h_t)
$.
The breaker-conditioned branch exploits symmetry after accounting for
hidden global context, while the raw branch handles localized observable
violations. Showing that this combined architecture outperforms both
single-view variants on mixed-breaker tasks is a primary empirical claim.\\
\noindent{\bf Evaluation.}
We will evaluate three regimes designed to isolate the two failure modes and their interaction. First, multi-robot navigation with
fixed obstacles isolates agent-wise local breakers. Second, hidden-wind navigation removes obstacles but introduces a per-episode wind vector omitted from observations, isolating latent global breakers. Third, multi-arm manipulation on an inclined surface with localized obstacles combines hidden gravity with local observed violations. Metrics include final return, sample efficiency, training stability across seeds, and view-specific diagnostics such as learned gates, latent-breaker estimation error, counterfactual equivariance error, etc.

% Metrics follow standard practice—final return, sample efficiency, training stability across seeds—plus view-specific diagnostics: gate activation maps and their correlation with oracle breaker labels (View~1), and angular error between $\widehat{b}_t$ and the true wind/gravity vector (View~2). Performance is reported as a function of symmetry-breaking strength to characterize the regime where each view, and the combined architecture, dominates.

% \textit{Baselines and ablations.} On View~1 we compare against a raw MLP critic, a fully invariant critic, and a \emph{current-breaker oracle gate} using ground-truth $\mathbf{1}\{(s,a) \in \mathcal{B}\}$—a diagnostic for whether current breaker information alone suffices (the future-exposure bound predicts it does not when trajectories cross breaker regions several steps ahead). On View~2: an unstructured MLP, a strict observation-only equivariant model, a recurrent non-equivariant model with history but no symmetry prior, an \emph{oracle} receiving the true $b$ (upper bound), and an \emph{observable-breaker ablation} exposing $b$ in the observation (confirming the latent variable is the missing information). We ultimately will compare the combined architecture against the strongest single-view variant from each.


\section{Expected results: key outcomes and timeline}
The project spans 14 months. The first three months focus on setup and baselines implementation. View 1 theory and algorithms are developed in months 2–6, running in parallel with View 2 theory and algorithms in months 4–8. Unification of the two views begins in month 8 and continues through month 11. The final three months are dedicated to code release, paper writing, and submission.

\section{Funds needed}
% \noindent{\bf Funds needed.}
\noindent{\em Cash funding:} \$42,450 is requested for the annual stipend of a research assistant (RA)~\citep{wpi_stipend_rates}.
\noindent{\em AWS promotional credits:}
\$50,000 is requested.
$\sim$\$35,000 will be allocated to EC2 for parallelized multi-robot RL training,
$\sim$\$10,000 to RoboMaker and SageMaker to enable realistic and large-scale robotics simulation, training, and benchmarking,
$\sim$\$5,000 to supporting infrastructure.


\bibliographystyle{unsrtnat}
\bibliography{references}

\end{document}
